---
title: "Balance Mutations in Logical Modeling"
author: "[John Zobolas](https://github.com/bblodfon)"
date: "Last updated: `r format(Sys.time(), '%d %B, %Y')`"
description: "Investigations for Balance link operators paper"
url: 'https\://bblodfon.github.io/balance-paper/'
github-repo: "bblodfon/balance-paper"
bibliography: references.bib
link-citations: true
site: bookdown::bookdown_site
---

# Intro {-}

Several analyses/investigations relating to the balance logical operators paper.

# Input {-}

Loading libraries:
```{r Load libraries, message = FALSE}
library(xfun)
library(knitr)
library(dplyr)
library(tidyr)
library(ggpubr)
```

# Analysis {-}

## Link Operators {-}

Let $f$ be a boolean function $f(x,y):\{0,1\}^n \rightarrow \{0,1\}$, with $m$ **activators** $x=\{x_i\}_{i=1}^{m}$ and $k$ **inhibitors** $y=\{y_i\}_{i=1}^{k}$, that is a total of $n=m+k$ regulators.
For a link operator to make sense, we have that $m,k \ge 1$ (at least one regulator in each category).
We next describe what each link operator looks like in the general form of $f$:

- `AND-NOT`: $$f(x,y) = \left(\bigvee_{i=1}^{m} x_i\right) \land \lnot \left(\bigvee_{i=1}^{k} y_i\right)$$
- `OR-NOT`: $$f(x,y) = \left(\bigvee_{i=1}^{m} x_i\right) \lor \lnot \left(\bigvee_{i=1}^{k} y_i\right)$$
- `BalanceOp1`: $$f(x,y) = \bigvee_{\forall (i,j)}^{m,k}(x_i\land \lnot y_j) = \left(\bigvee_{i=1}^{m} x_i\right) \land \left(\bigvee_{i=1}^{k} \lnot y_i\right)$$
- `exp_act_win`: $$f_{act-win}(x,y)=\begin{cases}
        1, & \text{for } \sum_{i=1}^{m} x_i \ge \sum_{i=1}^{k} y_i, \text{ with } \sum_{i=1}^{m} x_i \ne 0\\
        0, & \text{otherwise}
        \end{cases}$$
- `exp_inh_win`: $$f_{inh-win}(x,y)=\begin{cases}
        1, & \text{for } \sum_{i=1}^{m} x_i \gt \sum_{i=1}^{k} y_i, \text{ with } \sum_{i=1}^{m} x_i \ne 0\\
        0, & \text{otherwise}
        \end{cases}$$

Note that: $f_{inh-win}(x,y) = \lnot f_{act-win}(y,x)$.
I searched for an analytical formula for the two last functions (they get pertty big!).
See discussion on [math.stackexchange](https://math.stackexchange.com/questions/3767774/identify-boolean-function-that-satisfies-some-constrains/).

# Truth Density Data Analysis {-}

:::{.blue-box}
*Truth Density* of a boolean equation/expression, given it's equivalent truth table, is the **number of rows that the expression is active** divided to **the total number of rows** $(2^n)$.
:::

I created every possible truth table for up to $20$ variables (variables here means *regulators* for us) and calculated the `AND-NOT`, `OR-NOT`, `BalanceOp1`, `exp_act_win`, `exp_inh_win` results for every possible configuration of the number of activators and inhibitors that added up to the number of regulators.
Then, from the truth tables I calculated the **truth density** of each operator in each particular configuration.
See part of the data below:
```{r load-data}
stats = readRDS(file = "stats.rds")
stats[1:5,1:6] %>% kable(caption = "Thuth Density Data", digits = 2)
```

:::{.orange-box}
Use the [fun.R](https://github.com/bblodfon/balance-paper/blob/master/fun.R) script to reproduce this data.
:::

Also, I have proved the exact formulas for the truth densities in the case of the `AND-NOT` and `OR-NOT` link operators (see [here](http://tiny.cc/link-proofs) for a proof sketch).
I write them here explicitly, as well as their long-term behaviour (for large $n$. number of regulators):

- `AND-NOT`: $$TD_{AND-NOT}=\frac{2^m-1}{2^n} \xrightarrow{n \text{ large}} \frac{1}{2^k}$$
- `OR-NOT`:  $$TD_{OR-NOT}=\frac{2^n-2^k}{2^n} \xrightarrow{n \text{ large}} 1-\frac{1}{2^m}$$

So, if there are a lot of regulators in a boolean equation and **only one of them is an inhibitor (activator)** the truth density of the `AND-NOT` (`OR-NOT`) equation case will converge to $1/2$.

We can use the data above to validate the formulas from the proof (up to $n=20$):
```{r validate-formulas, comment=""}
# Validate AND-NOT Truth Density formula
formula_td_and_not = stats %>% 
  mutate(formula_td_and_not = (2^num_act - 1)/(2^num_reg)) %>%
  pull(formula_td_and_not)

all(stats %>% pull(td_and_not) == formula_td_and_not)

# Validate OR-NOT Truth Density formula
formula_td_or_not = stats %>% 
  mutate(formula_td_or_not = (((2^num_act - 1) * (2^num_inh)) + 1)/(2^num_reg)) %>%
  pull(formula_td_or_not)

all(stats %>% pull(td_or_not) == formula_td_or_not)
```

Comparing the `AND-NOT` and `OR-NOT` truth densities across the number of regulators:
```{r fig-and-or-not, fig.align='center', cache=TRUE, fig.cap="AND-NOT vs OR-NOT Truth Densities across all possible activators and inhibitors combinations up to 20 regulators"}
# tidy up data
stats_and_or = pivot_longer(data = stats, cols = c(td_and_not, td_or_not), 
  names_to = "lo", values_to = "td") %>%
  select(num_reg, lo, td) %>%
  mutate(lo = replace(x = lo, list = lo == "td_and_not", values = "AND-NOT")) %>%
  mutate(lo = replace(x = lo, list = lo == "td_or_not", values = "OR-NOT")) %>%
  rename(`Link Operator` = lo)

ggboxplot(data = stats_and_or, x = "num_reg", y = "td", 
  color = "Link Operator", palette = "Set1",
  title = "AND-NOT vs OR-NOT Truth Densities", 
  xlab = "Number of regulators", ylab = "Truth Density") +
  theme(plot.title = element_text(hjust = 0.5))
```

:::{.green-box}
- **The more regulators** there are, the more likely it is that the `AND-NOT` link operator in the boolean equation will result in an **inhibited** target and that the `OR-NOT` link operator in an **active** target.
- For $n>6$, the points outside the boxplots (with a truth density of $\frac{1}{2}, \frac{1}{4}, 1-\frac{1}{4},\frac{1}{8},1-\frac{1}{8},...$) correspond to the **long-term behaviour** of the truth density formulas shown above, where there is **large imbalance between the number of activators and inhibitors**.
:::


```{r}
# ggboxplot(data = stats, x = "num_reg", y = "td_balance_op")
# ggboxplot(data = stats, x = "num_reg", y = "td_exp_act")
# ggboxplot(data = stats, x = "num_reg", y = "td_exp_inh")
```


# R session info {-}

```{r session info, comment=""}
xfun::session_info()
```

# References {-}
